# Establishments

```{r}
#| label: setup
#| include: false

source("./scripts/setup.R")

data_label_size <- 3.5 # for plots using labels as data points

# for helpful sub-setting
top_naics_cutoff <- 5
top_county_cutoff <- 10

source("./scripts/define-reference-data.R")
source("./scripts/define-state-and-county-data.R")
source("./scripts/define-dataframes.R")

```

## Introduction

The County Business Patterns data uses the general term *establishment* to refer to corporations, partnerships, non-profit organizations, etc.

The CBP data includes establishments categorized by employee size class. The range is embedded in the name. For example, `n1_4` includes a count of establishments with 1-4 employees. The category `n1000` includes establishments with 1,000 or more employees.

The plots below that reference sectors use the sector abbreviations listed in @sec-tables.

<br>

## Number of establishments in NC

As one might expect, there are many more establishments with a small number of employees. Each subsequent employee size class includes fewer establishments. Compare the number of workers in each employee size class in @fig-state-workers.

```{r fig.height=6, fig.width=12}
#| label: fig-state-establishments-two-plot
#| fig-height: 6
#| fig-width: 12
#| fig-cap: "NC establishments 2021"
#| column: page-inset-right

p1 <- d_est_state_nc_total_size_classes |> 
  filter(year == 2021) |>
  ggplot() +
  geom_col(aes(x = value, y = est_size_class),
           fill = carolina_blue, alpha = 0.6) + 
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_discrete(limits=rev) +
  labs(
    x = "Establishments",
    y = "Employee size class",
    color = NULL,
  )

p2 <- d_est_state_nc_total_size_classes |> 
  mutate(est_size_class = as.character(est_size_class)) |>
  filter(year == 2021) |>
  inner_join(n_employees_ref,
             by = "est_size_class") |> 
  mutate(est_size_class = fct_reorder(est_size_class, pct_est)) |>
  arrange(desc(est_size_class)) |>
  mutate(cum_pct = cumsum(pct_est),
         ybar_start = lag(cum_pct, default = 0)) |>
  ggplot() +
  geom_segment(aes(x = est_size_class, xend = est_size_class, y = ybar_start, yend = cum_pct),
           color = carolina_blue, linewidth = 8, alpha = 0.6) + 
  scale_y_continuous(labels = label_percent()) +
  scale_x_discrete(limits=rev) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    x = "Employee size class",
    y = "Percent of establishments",
    color = NULL,
  )

p1 + p2 +
  plot_annotation(
    title = "NC establishments in each employee size class",
    subtitle = "2021; includes all sectors reported in CBP",
    caption = my_caption
  )

```

<br>

NC's distribution of establishments is quite similar to the USA as a whole:

```{r fig.height=6, fig.width=8}
#| label: fig-state-establishments-by-sector-pct-usa-compare
#| fig-height: 5
#| fig-width: 7
#| fig-cap: "NC establishments by sector compared to USA 2021"

nc_naics_abbr_levels <- d_est_sector_total_state_nc |>
  filter(year == 2021) |>
  arrange(pct_est) |>
  pull(naics_abbr)

data_for_plot <- d_est_sector_total_state_nc |>
  filter(year == 2021) |>
  mutate(scope = "nc") |>
  bind_rows(d_est_sector_total_usa |>
              filter(year == 2021) |>
              mutate(scope = "usa")
  ) |>
  select(-c(naics_abbr_num)) |>
  mutate(naics_abbr = factor(naics_abbr, levels = nc_naics_abbr_levels)) |>
  mutate(low_value = min(pct_est),
         high_value = max(pct_est),
         .by = naics_abbr) |>
  mutate(diff_pct = high_value - low_value,
         diff_pct = if_else(pct_est == high_value, diff_pct, -1 * diff_pct)) 

diff_pct_cutoff <- 0.005

data_for_plot |>
  ggplot() +
  geom_segment(aes(x = low_value, xend = high_value, y = naics_abbr, yend = naics_abbr),
               alpha = 0.3) +
  geom_point(aes(x = pct_est, y = naics_abbr, color = scope),
           alpha = 0.6) + 
  geom_point(aes(x = pct_est, y = naics_abbr, color = scope),
           alpha = 0.6) + 
  geom_text(aes(x = high_value, y = naics_abbr,
                label = if_else(diff_pct > diff_pct_cutoff,
                                percent(pct_est, accuracy = .1),
                                NA_character_)),
            na.rm = TRUE, size = 3,
            hjust = 0, nudge_x = 0.003) +
  geom_text(aes(x = low_value, y = naics_abbr,
                label = if_else(diff_pct < -1 * diff_pct_cutoff,
                                percent(pct_est, accuracy = .1),
                                NA_character_)),
            na.rm = TRUE, size = 3,
            hjust = 1, nudge_x = -0.003) +
  scale_x_continuous(labels = label_percent()) +
  scale_color_manual(values = c(carolina_blue, "firebrick")) +
  theme(legend.position = c(0.8, 0.4),
        legend.box.background = element_rect(fill = "grey80")) +
  labs(
    title = "NC establishments by sector\ncompared to USA 2021",
    subtitle = glue("Includes all 100 counties and 'Statewide'", 
                    "; showing percentages where difference is > {percent(diff_pct_cutoff, accuracy = 0.1)}"),
    x = "Establishements",
    y = "",
    color = NULL,
    caption = my_caption
  )

```

<br>

## Trends in the number of establishements

The 2008 financial crisis led to a significant drop in the number of construction, real estate, and retail establishments. In most sectors, establishments were able to shed employees during the crisis and stay in business (compare @fig-state-employment-naics-timeseries), presumably due in part to the Paycheck Protection Program and other assistance from Federal^[<https://home.treasury.gov/policy-issues/coronavirus/assistance-for-small-businesses> ] and state^[<https://www.ncdor.gov/business-recovery-grant> ] governments. 

```{r fig.height=10, fig.width=12}
#| label: fig-state-establishments-naics-timeseries
#| fig-height: 10
#| fig-width: 12
#| fig-cap: "NC establishments in each sector 2001-2021"
#| column: page-inset-right
#| warning: false

data_for_plot <- d_est_sector_total_state_nc |>
  filter(naics_abbr != "OtherInd") |>
  mutate(naics_abbr = fct_reorder(naics_abbr, -est)
         ) |>
  mutate(est_fist_year = est[year == min(year)],
         est_last_year = est[year == max(year)],
         .by = c("naics")) |>
  mutate(pct_est_diff = est_last_year / est_fist_year - 1)

labels_for_plot <- data_for_plot |>
  filter(year == max(year))

p1 <- data_for_plot |>
  filter(naics_abbr != "Other") |>
  ggplot() +
  geom_line(aes(x = year, y = est, color = naics_abbr, group = naics_abbr),
           linewidth = 0.4, alpha = 0.6, show.legend = FALSE) +
  geom_text_repel(data = labels_for_plot,
                  aes(x = year + 0.5, y = est, label = naics_abbr, color = naics_abbr),
                  direction = "y", hjust = 0, vjust = 0.5, size = 3,
                  min.segment.length = unit(1, "cm"), force_pull = 100,
                  seed = 123,
                  show.legend = FALSE) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  expand_limits(x = 2026,
                y = 0) +
  guides(color = guide_legend(override.aes = c(linewidth = 3))) +
  labs(
    subtitle = "NC establishments in each sector",
    x = "",
    y = "Number of establishments",
    color = NULL,
    caption = my_caption
  )

p2 <- data_for_plot |>
  filter(naics_abbr != "Other") |>
  ggplot() +
  geom_area(aes(x = year, y = est, color = naics_abbr, fill = naics_abbr, group = naics_abbr),
           linewidth = 0.4, alpha = 0.6, show.legend = FALSE) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                     expand = expansion(mult = c(0, 0.02))) +
  expand_limits(#x = 2026,
                y = 0) +
  guides(color = guide_legend(override.aes = c(linewidth = 3))) +
  labs(
    subtitle = "NC establishments combined trend",
    x = "",
    y = "Number of establishments",
    color = NULL,
    caption = my_caption
  )

p3 <- data_for_plot |>
  filter(year == 2021,
         est != 0) |>
  ggplot(aes(est, pct_est_diff, color = naics_abbr)) +
  geom_hline(yintercept = 0, lty = 2, size = 0.2, alpha = 0.2) +
  geom_label(aes(label = naics_abbr),
             show.legend = FALSE, hjust = 0.5, vjust = 0.5) +
  scale_x_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    subtitle = glue("Percent change in establishments since 2001"),
    x = "Establishments in NC in 2021",
    y = "Pct change since 2001"
    )

(p1 + p2) / p3 +
  plot_annotation(
    title = "NC establishments in each sector 2001-2021",
    subtitle = "Includes all sectors reported in CBP",
    caption = my_caption
  )

```


<br>

How has the proportion of establishments by employee size class changed over time? In general, the smallest establishments come and go faster than larger establishments.

```{r fig.height=8, fig.width=10}
#| label: fig-state-establishments-pct-timeseries
#| fig-height: 8
#| fig-width: 10
#| fig-cap: "Percent of NC establishments by employee size class 2001-2021 for top sectors"
#| column: page-inset-right

data_for_plot <- d_est_state_nc_size_classes |>
  mutate(naics_abbr = fct_lump(naics_abbr, 9, w = value)) |>
  filter(naics_abbr != "Other")

data_for_plot |>
  ggplot() +
  geom_line(aes(x = year, y = pct_est, color = est_size_class, group = est_size_class),
           linewidth = 0.4, alpha = 0.6) +
  scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
  facet_wrap(. ~ naics_abbr, scales = "free_y") +
  guides(color = guide_legend(override.aes = c(linewidth = 3))) +
  labs(
    title = "Percent of NC establishments in each employee size class\nby sector 2001-2021",
    subtitle = "Top 9 sectors by number of establishments (in 2021) as reported in CBP",
    x = "",
    y = "Percent of establishments (scale varies)",
    color = NULL,
    caption = my_caption
  )

```

<br>

## Table

Sorted by percent difference in number of 2001 to 2021.

```{r}
#| label: tbl-state-nc-est
#| tbl-cap: "NC state establishments by sector 2021"
#| tbl-cap-location: bottom

d_est_sector_total_state_nc |>
  filter(naics_abbr != "OtherInd") |>
  mutate(naics_abbr = fct_reorder(naics_abbr, -est)
         ) |>
  mutate(est_fist_year = est[year == min(year)],
         est_last_year = est[year == max(year)],
         .by = c("naics")) |>
  mutate(pct_est_diff = est_last_year / est_fist_year - 1) |>
  filter(year == 2021) |>
  select(naics, naics_abbr, est_2021 = est, pct_est_2021 = pct_est, pct_est_diff_since_2001 = pct_est_diff) |>
  arrange(-pct_est_diff_since_2001) |>
  mutate(rowid = row_number()) |>
  gt() |>
  tab_header(md(glue("**NC sector establishments in 2021 and growth since 2001**",
                     "<br>*Sorted by growth since 2001*"))) |>
  tab_source_note(md("*US Census County Business Patterns; analysis by Daniel Moul*")) |>
  tab_options(table.font.size = 10) |>
  fmt_number(columns = c(est_2021),
             decimals = 0) |>
  fmt_percent(columns = c(pct_est_2021, pct_est_diff_since_2001),
             decimals = 0)
```

<br>
